---
title: "Package Skeleton"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package Skeleton}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor:
  markdown:
    wrap: sentence
---

# Motivation

@cpp4r is better used when adding C++ code to R packages, as it allows for proper script organization and documentation. The reference for this vignette is @usethis.

# Loading the Required R Packages

This vignette and the next use the following R packages:

```r
library(cpp4r)
library(devtools)
library(bench)
library(usethis)
```

# Running a single function

It is possible to run a single function in C++ using the `cpp4r::cpp_source()` function. For example, to create a function that adds one to a number, you can run:

```{r}
cpp4r::cpp_source(
  code = "
    #include <cpp4r.hpp>

    using namespace cpp4r;

    [[cpp4r::register]] int plusone(int x) {
      return x + 1;
    }",
  quiet = FALSE
)
```

This function can be called from R by running `plusone(2)`, which should return 3, or with any other integer.

However, it is a much better practice to organize C++ code in an R package.

# Creating a Dummy Package

You can create a new package in RStudio (or VSCode) by running:

```r
create_package("~/github/cpp-for-r-users/cpp4rexamples")
```

This command will create a new folder named `cpp4rexamples`. Afterward, you can run `use_cpp4r()` to add the necessary files for using C++ code within R.

# Print a Number

You can create a new R script by running `use_r("cpp4rexamples-package")`. You should add the following code to the file:

```r
#' @useDynLib cpp4rexamples, .registration = TRUE
NULL
```

The `usethis` package will automatically create the file `src/code.cpp`. You can modify it to include a simple function that prints the number 1:

```cpp
#include "cpp4r.hpp"

using namespace cpp4r;

[[cpp4r::register]] int one_cpp_() { 
  return 1; 
}
```

Within R, there is no need to create a `main()` function in C++.

The R version of the function is as follows:

```r
#' Return 1 (R)
#' @export
one_r <- function() {
    1L
}
```

To export the C++ function, you can add the following lines to `cpp4rexamples-package.R`, which documents the function and allows for defining default argument values:

```r
#' Return 1 (C++)
#' @export
one_cpp <- function() {
  one_cpp_()
}
```

# Number's Sign

A more complex function is one that returns the sign of a number:

```cpp
[[cpp4r::register]] int sign_cpp_(double x) {
  if (x > 0) {
    return 1;
  } else if (x == 0) {
    return 0;
  } else {
    return -1;
  }
}
```

Here is the R version of the function:

```r
#' Return the sign of a number (R)
#' @param x integer
#' @export
sign_r <- function(x) {
  if (x > 0) {
    1
  } else if (x == 0) {
    0
  } else {
    -1
  }
}
```

You should also add the corresponding auxiliary function for documentation:

```r
#' Return the sign of a number (C++)
#' @inheritParams sum_r
#' @export
sign_cpp <- function(x) {
  sign_cpp_(x)
}
```

# Installing the Package

To document and install the package as an R library, you can run the following functions:

```r
cpp_register()
document()
install()
```

Afterward, you can access the functions by loading the package with `library(cpp4rexamples)`.

For development and live testing, you can use the `load_all()` function:

```{r pkg-load}
load_all()
```

Each time you need to make changes to the C++ code, you can run `load_all()` again to test and then reinstall the package.

# Good Practice

It is good practice to include a license for your code. For example, you can use the Apache license by running:

```r
use_apache_license()
```

Additionally, it is recommended to use `use_build_ignore()` to ignore files that are unnecessary for package installation. For example, to ignore the `docs` folder, you can run:

```r
use_build_ignore("docs")
```

Your `.Rbuildignore` file can include the following lines, which you can edit manually to exclude specific files and directories:

```
^\.vscode$
^LICENSE\.md$
^cpp4rexamples\.Rproj$
```

# References
