---
title: "Logical Functions with Missing Values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Logical Functions with Missing Values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor:
  markdown:
    wrap: sentence
---

# Motivation

This vignette expands vignette 6 by using the same methods from vignette 6. These examples were adapted from @vaughan.

R already provides efficient versions of the functions covered here. Code optimizations and improvements will be made in later vignettes.

# Load the Package

You can load the `cpp4rexamples` package after adding the functions from the next sections to it with the following code:

```{r 03-setup}
load_all()
```

# Additional Packages

You can use the `bench` package to compare the performance of the functions. You can load the package with the following code:

```{r 03-bench}
library(bench)
```

# Are Some Values True? (`any()`)

The `any()` function returns `TRUE` if there is at least one `TRUE` element in a vector, and `FALSE` otherwise. Below is one possible C++ implementation:

```cpp
[[cpp4r::register]] bool any2_cpp_(logicals x, bool na_rm = false) {
  int n = x.size();
  bool has_na = false;
  
  for (int i = 0; i < n; ++i) {
    if (x[i] == NA_LOGICAL) {
      has_na = true;
      if (!na_rm) {
        continue; // Skip NA values if na_rm is false
      }
    } else if (x[i]) {
      return true;
    }
  }
  
  // If we found any NA and na_rm is false, return NA
  if (has_na && !na_rm) {
    return NA_LOGICAL;
  }
  
  return false;
}
```

Unlike the `any()` function from vignette 6, this function has an additional argument `na_rm` that allows the user to remove missing values from the consideration.

The corresponding auxiliary function for documentation is:

```r
#' Return TRUE if any element in a vector is TRUE (C++)
#' @param x logical vector
#' @param na_rm logical. Should missing values (including `NaN`) be removed?
#' @export
any2_cpp <- function(x, na_rm = FALSE) {
  any2_cpp_(x, na_rm = na_rm)
}
```

To test the functions, you can run the following benchmark code in the R console:

```{r 04-functions-bench2}
set.seed(123) # for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements
y <- ifelse(x > 2, TRUE, FALSE)

any(c(TRUE, NA, FALSE))
any2_cpp(c(TRUE, NA, FALSE))
any2_cpp(c(TRUE, NA, FALSE), na_rm = TRUE)

mark(
  any(y),
  any2_cpp(y)
)
```

# Which Indices are TRUE? (`which()`)

The `which()` function returns the indices of the `TRUE` elements in a vector. Here is a possible C++ implementation:

```cpp
[[cpp4r::register]] integers which2_cpp_(logicals x, bool na_rm = false) {
  int n = x.size();
  writable::integers res;
  int j = 0;

  for (int i = 0; i < n; ++i) {
    if (x[i] == NA_LOGICAL) {
      if (!na_rm) {
        // Skip NA values if na_rm is false
        continue;
      }
    } else if (x[i]) {
      ++j;
      res.push_back(i + 1);
    }
  }

  if (j == 0) {
    return integers(0);
  } else {
    return res;
  }
}
```

The corresponding auxiliary function for documentation is:

```r
#' Return the index of the TRUE elements in a vector (C++)
#' @param x logical vector
#' @param na_rm logical. Should missing values (including `NaN`) be removed?
#' @export
which2_cpp <- function(x, na_rm = FALSE) {
  which2_cpp_(x, na_rm = na_rm)
}
```

To test the functions, you can run the following benchmark code in the R console:

```{r 04-functions-which-bench}
which(c(TRUE, NA, FALSE, TRUE))
which2_cpp(c(TRUE, NA, FALSE, TRUE))
which2_cpp(c(TRUE, NA, FALSE, TRUE), na_rm = TRUE)

mark(
  which(y[1:1000]),
  which2_cpp(y[1:1000])
)
```

# Are All Values True? (`all()`)

The `all()` function checks if all elements in a vector are `TRUE`. Here is a possible C++ implementation that loops over the vector:

```cpp
[[cpp4r::register]] bool all2_cpp_1_(logicals x, bool na_rm = false) {
  int n = x.size();
  bool has_na = false;
  
  for (int i = 0; i < n; ++i) {
    if (x[i] == NA_LOGICAL) {
      has_na = true;
      if (!na_rm) {
        continue; // Skip NA values if na_rm is false
      }
    } else if (!x[i]) {
      return false;
    }
  }
  
  // If we found any NA and na_rm is false, return NA
  if (has_na && !na_rm) {
    return NA_LOGICAL;
  }
  
  return true;
}
```

More concise C++ alternatives are:

```cpp
[[cpp4r::register]] bool all2_cpp_2_(logicals x, bool na_rm = false) {
  bool has_na = false;
  
  for (int i = 0; i < x.size(); ++i) {
    if (x[i] == NA_LOGICAL) {
      has_na = true;
      if (!na_rm) {
        continue;
      }
    } else if (!x[i]) {
      return false;
    }
  }
  
  if (has_na && !na_rm) {
    return NA_LOGICAL;
  }
  
  return true;
}

[[cpp4r::register]] bool all2_cpp_3_(logicals x, bool na_rm = false) {
  bool has_na = false;
  
  for (int i = 0; i < x.size(); ++i) {
    if (x[i] == NA_LOGICAL) {
      has_na = true;
      if (!na_rm) {
        continue;
      }
    } else if (!x[i]) {
      return false;
    }
  }
  
  if (has_na && !na_rm) {
    return NA_LOGICAL;
  }
  
  return true;
}
```

The corresponding auxiliary function for documentation is:

```r
#' Return TRUE if all elements in a vector are TRUE (C++)
#' @param x logical vector
#' @param na_rm logical. Should missing values (including `NaN`) be removed?
#' @export
all2_cpp <- function(x, na_rm = FALSE) {
  all2_cpp_1_(x, na_rm = na_rm)
}
```

To avoid typing `std::` every time, you can use `using namespace std;` at the top of `src/code.cpp`. However, this is not recommended because it can lead to conflicts. A better option is to declare `using std::the_function;` which means you can use `the_function` instead of `std::the_function` each time [@akbiggs].

To test the functions, you can run the following tests and benchmark code in the R console:

```{r 05-functions-all-correctness}
set.seed(123) # for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

all(c(TRUE, NA, TRUE))
all2_cpp_1_(c(TRUE, NA, TRUE))
all2_cpp_1_(c(TRUE, NA, TRUE), na_rm = TRUE)

all(c(FALSE, NA, TRUE))
all2_cpp_1_(c(FALSE, NA, TRUE))
all2_cpp_1_(c(FALSE, NA, TRUE), na_rm = TRUE)

# also test the TRUE-only case
all(x >= 0)
all2_cpp_1_(x >= 0)
all2_cpp_2_(x >= 0)
all2_cpp_3_(x >= 0)
```

```{r 05-functions-all-benchmark}
mark(
  all(x > 2),
  all2_cpp_1_(x > 2),
  all2_cpp_2_(x > 2),
  all2_cpp_3_(x > 2)
)
```

# References
